Factor analysis of car miles per gallon (MPG)
==========================


This document is my Course Project submission for the class "Regression Models" on Coursera.

# Summary

This study is an analysis of the very famous <mtcars dataset>(https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html), available in R. It aims at explaining the possible relationships between a car's miles per gallon measure, and the different other characteristics of the car. In particular, we will focus our attention on the transmission in order to give an answer to the two following questions:

* Is an automatic or manual transmission better for MPG?
* How can we quantify the MPG difference between automatic and manual transmissions?

First of all, we will explore the data to get a better overview of the underlying patterns in the dataset, and then we will build several linear models to explain and quantify the relationships between transmission and MPG.

# Analysis

## Data exploration

### Loading data

```{r}
data(mtcars)
str(mtcars)
```

Our main focus is the relationship between transmission "am" variable (binary: 0 = automatic, 1 = manual) and the number of miles per US gallon "mpg" variable. For ease of use, let's build a "transmission" factor variable out of the "am"" variable:

```{r}
mtcars$transmission <- factor(mtcars$am, labels = c("automatic", "manual"))
```

### Data visualisation

Visualising "transmission" and "MPG" variables can help us understand the main patterns in the data. The figure 1 shows that manual cars tend to have higher MPG values. This is confirmed by Figure 2, as it appears clearly that the distribution of MPG for manual cars is higher than for automatic cars.

```{r}
library(ggplot2)
g1 <- ggplot(mtcars, aes(x=mpg, fill = transmission)) + 
  geom_histogram(binwidth = 1, col = "black") +
  ggtitle('Figure 1: Histogram of Miles per US Gallon values')
g2 <- ggplot(mtcars, aes(x=transmission, y=mpg, fill = transmission)) +
  geom_boxplot(adjust = 1) + geom_jitter(size = 2) +
  ggtitle('Figure 2: Boxplot of MPG values for each transmission')
```

Thus, this gives an indication about the answer of the first question: manual cars seems to be better for MPG. We should now try to model and quantify this relationship.

## Simple model


## Further models and quantitative analysis

# Appendix
```{r, echo=FALSE}
g1
g2
```
